---
title: "Mean residual life"
author: "Victor Rwandarwacu"
date: "2025-03-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```





```{r}
# Load required libraries
library(survival)
library(survminer)
library(dplyr)


library(KMsurv)

surv_brca_cli <- read.csv("surv_brca_cli.csv")
surv_brca_cli= surv_brca_cli %>% mutate(group= pathologic_stage) %>% mutate( group= case_when( group=="Stage I"~ "Group A",group=="Stage II"~ "Group B",group=="Stage III"~ "Group C",group=="Stage IV"~ "Group D",group=="Stage X"~ "Group E",group=="Unspecified"~ "Group F"))

```


## mean resuial life stage 2vs 3 

```{r}
library(survival)
library(ggplot2)
library(dplyr)

# Filter and prepare the data
filtered_data <- surv_brca_cli %>%
  filter(group %in% c("Group B","Group C")) %>%
  mutate(group = factor(group))

summary(filtered_data)
# survival Object
surv_object <- Surv(time = filtered_data$time_in_years, event = filtered_data$vital_status)

# KM fit
fit <- survfit(surv_object ~ group, data = filtered_data)
summary(fit)
# Function to compute MRL from survival fit
compute_mrl <- function(fit) {
  time <- fit$time
  surv_probs <- fit$surv
  strata <- rep(names(fit$strata), fit$strata)

  df <- data.frame(time = time, surv = surv_probs, strata = strata)

  mrl_df <- df %>%
    group_by(strata) %>%
    arrange(time) %>%
    mutate(
      delta_t = lead(time, default = max(time)) - time,
      area = surv * delta_t,
      mrl = rev(cumsum(rev(area)))  # Mean residual life
    )

  return(mrl_df)
}

# Compute MRL
mrl_data <- compute_mrl(fit)

# Plot Mean Residual Life
ggplot(mrl_data, aes(x = time, y = mrl, color = strata)) +
  geom_line(size = 1.2) +
  labs(
    title = "Mean Residual Life: Group A vs Group B",
    x = "Time in Years",
    y = "Mean Residual Life (Years)",
    color = "Group"
  ) +
  theme_minimal()+ 
  scale_color_manual(
    labels = c("Group A", "Group B"),
    values = c("blue", "red")  # or use your desired colors
  )


```

# interpretation 

At any given point in follow-up, a Stage II patient has more expected years of life remaining than a Stage III patient — reflecting the impact of cancer stage on survival expectations. The difference is greatest early in follow-up and diminishes as time progresses

consider the distance between the 2 plots . the curves converges around 12 to 16 years.

Patients with pathologic Stage II breast cancer consistently have better prognosis and longer expected survival, even after surviving up to a certain time.

## mean residual life stage 2 vs 4 

```{r}
# Load required libraries
library(survival)
library(ggplot2)
library(dplyr)

# Filter and prepare the data for Stage II and Stage IV
filtered_data <- surv_brca_cli %>%
  filter(pathologic_stage %in% c("Stage II", "Stage IV")) %>%
  mutate(pathologic_stage = factor(pathologic_stage))

# Create survival object
surv_object <- Surv(time = filtered_data$time_in_years, event = filtered_data$vital_status)

# Fit Kaplan-Meier survival model
fit <- survfit(surv_object ~ pathologic_stage, data = filtered_data)

# Function to compute Mean Residual Life from survfit object
compute_mrl <- function(fit) {
  time <- fit$time
  surv_probs <- fit$surv
  strata <- rep(names(fit$strata), fit$strata)

  df <- data.frame(time = time, surv = surv_probs, strata = strata)

  mrl_df <- df %>%
    group_by(strata) %>%
    arrange(time) %>%
    mutate(
      delta_t = lead(time, default = max(time)) - time,
      area = surv * delta_t,
      mrl = rev(cumsum(rev(area)))  # Mean Residual Life
    )

  return(mrl_df)
}

# Compute MRL values
mrl_data <- compute_mrl(fit)

# Plot Mean Residual Life
ggplot(mrl_data, aes(x = time, y = mrl, color = strata)) +
  geom_line(size = 1.2) +
  labs(
    title = "Mean Residual Life: Group A vs Group B",
    x = "Time in Years",
    y = "Mean Residual Life (Years)",
    color = "Group"
  ) +
  theme_minimal()+ 
  scale_color_manual(
    labels = c("Group A", "Group B"),
    values = c("blue", "red")  # or use your desired colors
  )

```
##Interpration 

the stage 4 curves drops quickls and ends


The plot clearly illustrates that Stage IV cancer is associated with drastically worse survival expectations than Stage II — both at diagnosis and at every point in time thereafter. Stage II patients have more life years to expect and the decline in their MRL is more gradual, reflecting more favorable long-term survival.

## interporation of the difference plot 
```{r}
# Load libraries
library(survival)
library(ggplot2)
library(dplyr)

# Function to compute MRL from survfit
compute_mrl <- function(fit) {
  time <- fit$time
  surv_probs <- fit$surv
  strata <- rep(names(fit$strata), fit$strata)

  df <- data.frame(time = time, surv = surv_probs, strata = strata)

  mrl_df <- df %>%
    group_by(strata) %>%
    arrange(time) %>%
    mutate(
      delta_t = lead(time, default = max(time)) - time,
      area = surv * delta_t,
      mrl = rev(cumsum(rev(area)))
    )
  return(mrl_df)
}

# Function to interpolate MRL values on common time grid
interpolate_mrl <- function(mrl_df, group_name, grid_times) {
  approx_df <- approx(mrl_df$time, mrl_df$mrl, xout = grid_times, rule = 2)
  data.frame(
    time = grid_times,
    mrl = approx_df$y,
    group = group_name
  )
}

# Get MRL data for 2 vs 3
data_2_3 <- surv_brca_cli %>%
  filter(pathologic_stage %in% c("Stage II", "Stage III")) %>%
  mutate(pathologic_stage = factor(pathologic_stage))

fit_2_3 <- survfit(Surv(time_in_years, vital_status) ~ pathologic_stage, data = data_2_3)
mrl_2_3 <- compute_mrl(fit_2_3)

# Get MRL data for 2 vs 4
data_2_4 <- surv_brca_cli %>%
  filter(pathologic_stage %in% c("Stage II", "Stage IV")) %>%
  mutate(pathologic_stage = factor(pathologic_stage))

fit_2_4 <- survfit(Surv(time_in_years, vital_status) ~ pathologic_stage, data = data_2_4)
mrl_2_4 <- compute_mrl(fit_2_4)

# Create a common time grid
common_times <- sort(unique(c(mrl_2_3$time, mrl_2_4$time)))
common_times <- common_times[common_times > 0]  # Avoid time zero issues

# Interpolate both comparisons on common time grid
mrl_2_3_stage2 <- interpolate_mrl(filter(mrl_2_3, strata == "pathologic_stage=Stage II"), "Stage II", common_times)
mrl_2_3_stage3 <- interpolate_mrl(filter(mrl_2_3, strata == "pathologic_stage=Stage III"), "Stage III", common_times)

mrl_2_4_stage2 <- interpolate_mrl(filter(mrl_2_4, strata == "pathologic_stage=Stage II"), "Stage II", common_times)
mrl_2_4_stage4 <- interpolate_mrl(filter(mrl_2_4, strata == "pathologic_stage=Stage IV"), "Stage IV", common_times)

# Compute MRL differences
mrl_diff_2_3 <- data.frame(
  time = common_times,
  mrl_diff = mrl_2_3_stage2$mrl - mrl_2_3_stage3$mrl,
  comparison = "Stage II - Stage III"
)

mrl_diff_2_4 <- data.frame(
  time = common_times,
  mrl_diff = mrl_2_4_stage2$mrl - mrl_2_4_stage4$mrl,
  comparison = "Stage II - Stage IV"
)


```




## no extrapolation

```{r}
# Determine max time for shortest group in Stage II vs III
max_time_2_3 <- min(
  max(mrl_2_3_stage2$time, na.rm = TRUE),
  max(mrl_2_3_stage3$time, na.rm = TRUE)
)

# Plot with x-axis limited
ggplot(mrl_diff_2_3 %>% filter(time <= max_time_2_3), aes(x = time, y = mrl_diff)) +
  geom_line(color = "steelblue", size = 1.2) +
  labs(
    title = "Mean Residual Life Difference: Group A vs B",
    x = "Time in Years",
    y = "MRL Difference (Group A - B)"
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  xlim(0, max_time_2_3) +
  theme_minimal()


```




```{r}
# Determine max time for shortest group in Stage II vs IV
max_time_2_4 <- min(
  max(mrl_2_4_stage2$time, na.rm = TRUE),
  max(mrl_2_4_stage4$time, na.rm = TRUE)
)

# Plot with x-axis limited
ggplot(mrl_diff_2_4 %>% filter(time <= max_time_2_4), aes(x = time, y = mrl_diff)) +
  geom_line(color = "firebrick", size = 1.2) +
  labs(
    title = "Mean Residual Life Difference: Group A vs B",
    x = "Time in Years",
    y = "MRL Difference (Group A - B)"
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  xlim(0, max_time_2_4) +
  theme_minimal()
```
```{r}
# Load necessary libraries
library(survival)
library(ggplot2)
library(dplyr)
library(tidyr)

# Load your data
surv_brca_cli <- read.csv("surv_brca_cli.csv")

# Utility: compute MRL
compute_mrl <- function(fit) {
  time <- fit$time
  surv_probs <- fit$surv
  strata <- rep(names(fit$strata), fit$strata)
  
  df <- data.frame(time = time, surv = surv_probs, strata = strata)
  
  df %>%
    group_by(strata) %>%
    arrange(time) %>%
    mutate(
      delta_t = lead(time, default = max(time)) - time,
      area = surv * delta_t,
      mrl = rev(cumsum(rev(area)))
    ) %>%
    select(time, mrl, strata)
}

# Interpolate MRL to common grid
interpolate_mrl <- function(mrl_df, strata_label, grid) {
  mrl_interp <- approx(mrl_df$time, mrl_df$mrl, xout = grid, rule = 2)$y
  tibble(time = grid, mrl = mrl_interp, strata = strata_label)
}

# Bootstrap function
bootstrap_mrl_diff <- function(data, stage1, stage2, n_boot = 500) {
  set.seed(123)
  data <- data %>% filter(pathologic_stage %in% c(stage1, stage2))
  data$pathologic_stage <- factor(data$pathologic_stage)
  
  # Common time grid from original data
  surv_obj <- Surv(data$time_in_years, data$vital_status)
  fit <- survfit(surv_obj ~ pathologic_stage, data = data)
  mrl_orig <- compute_mrl(fit)
  common_time <- sort(unique(mrl_orig$time))
  
  # Original interpolated MRL
  mrl1 <- interpolate_mrl(filter(mrl_orig, strata == paste0("pathologic_stage=", stage1)), stage1, common_time)
  mrl2 <- interpolate_mrl(filter(mrl_orig, strata == paste0("pathologic_stage=", stage2)), stage2, common_time)
  mrl_diff_mean <- mrl1$mrl - mrl2$mrl
  
  # Matrix to hold bootstrapped differences
  boot_matrix <- matrix(NA, nrow = length(common_time), ncol = n_boot)
  
  for (i in 1:n_boot) {
    boot_sample <- data %>% group_by(pathologic_stage) %>% slice_sample(prop = 1, replace = TRUE) %>% ungroup()
    fit_boot <- survfit(Surv(time_in_years, vital_status) ~ pathologic_stage, data = boot_sample)
    
    mrl_boot <- compute_mrl(fit_boot)
    
    try({
      mrl1_boot <- interpolate_mrl(filter(mrl_boot, strata == paste0("pathologic_stage=", stage1)), stage1, common_time)
      mrl2_boot <- interpolate_mrl(filter(mrl_boot, strata == paste0("pathologic_stage=", stage2)), stage2, common_time)
      boot_matrix[, i] <- mrl1_boot$mrl - mrl2_boot$mrl
    }, silent = TRUE)
  }
  
  # CI computation
  ci_lower <- apply(boot_matrix, 1, quantile, probs = 0.025, na.rm = TRUE)
  ci_upper <- apply(boot_matrix, 1, quantile, probs = 0.975, na.rm = TRUE)
  
  result <- tibble(
    time = common_time,
    diff = mrl_diff_mean,
    lower = ci_lower,
    upper = ci_upper
  )
  
  return(result)
}

# Run bootstrap for both comparisons
mrl_ci_2_3 <- bootstrap_mrl_diff(surv_brca_cli, "Stage II", "Stage III", n_boot = 500)
mrl_ci_2_4 <- bootstrap_mrl_diff(surv_brca_cli, "Stage II", "Stage IV", n_boot = 500)

# Plot: Stage II vs III
ggplot(mrl_ci_2_3, aes(x = time, y = diff)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.3) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.8) + 
  labs(
    title = "MRL Difference with 95% CI: Group A vs B",
    x = "Time in Years",
    y = "MRL Difference (A - B)_Years"
  ) +
  theme_minimal()

# Plot: Stage II vs IV
ggplot(mrl_ci_2_4, aes(x = time, y = diff)) +
  geom_line(color = "firebrick", size = 1.2) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "firebrick", alpha = 0.3) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.8) + 
  labs(
    title = "MRL Difference with 95% CI: Group A vs B",
    x = "Time in Years",
    y = "MRL Difference (Group A- B)_Years"
  ) +
  theme_minimal()

```
########################################################################################################
#Exact time point plot 
#################################################################################


```{r}
# STEP 1: Load required packages
library(survival)
library(dplyr)

# STEP 2: Filter your data
df <- surv_brca_cli %>%
  filter(pathologic_stage %in% c("Stage II", "Stage IV")) %>%
  mutate(pathologic_stage = factor(pathologic_stage))
summary(df)
# STEP 3: Fit KM model
fit <- survfit(Surv(time_in_years, vital_status) ~ pathologic_stage, data = df)

# STEP 4: Extract MRL for each group
compute_mrl <- function(fit) {
  time <- fit$time
  surv_probs <- fit$surv
  strata <- rep(names(fit$strata), fit$strata)
  
  df <- data.frame(time = time, surv = surv_probs, strata = strata) %>%
    group_by(strata) %>%
    arrange(time) %>%
    mutate(
      delta_t = lead(time, default = max(time)) - time,
      area = surv * delta_t,
      mrl = rev(cumsum(rev(area)))
    )
  return(df %>% select(time, mrl, strata))
}

mrl_df <- compute_mrl(fit)

# STEP 5: Split by group
mrl_ii <- filter(mrl_df, strata == "pathologic_stage=Stage II")
mrl_iii <- filter(mrl_df, strata == "pathologic_stage=Stage IV")

# STEP 6: Determine shared time grid
max_common_time <- min(max(mrl_ii$time), max(mrl_iii$time))
shared_time <- sort(unique(c(mrl_ii$time, mrl_iii$time)))
shared_time <- shared_time[shared_time <= max_common_time]

# STEP 7: Use step function to evaluate MRL at shared times
step_mrl_ii <- stepfun(mrl_ii$time, c(mrl_ii$mrl, tail(mrl_ii$mrl, 1)))
step_mrl_iii <- stepfun(mrl_iii$time, c(mrl_iii$mrl, tail(mrl_iii$mrl, 1)))

# STEP 8: Create aligned data frame
aligned_df <- data.frame(
  time = shared_time,
  MRL_II = step_mrl_ii(shared_time),
  MRL_III = step_mrl_iii(shared_time),
  MRL_Diff = step_mrl_ii(shared_time) - step_mrl_iii(shared_time)
)

# STEP 9: View or plot the result
head(aligned_df)

```
```{r}
library(ggplot2)

ggplot(aligned_df, aes(x = time, y = MRL_Diff)) +
  geom_step(color = "steelblue", linewidth = 1.2) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "MRL Difference: Stage II vs Stage III ",
    x = "Time (Years)",
    y = "Mean Residual Life Difference"
  ) +
  theme_minimal()

```
## WIth confidence interval calculated using Delta method for variance 
```{r}
# Required libraries
library(survival)
library(dplyr)
library(ggplot2)

# Filter data
df <- surv_brca_cli %>%
  filter(pathologic_stage %in% c("Stage II", "Stage III")) %>%
  mutate(pathologic_stage = factor(pathologic_stage))

# Fit Kaplan-Meier model
fit <- survfit(Surv(time_in_years, vital_status) ~ pathologic_stage, data = df)

# Greenwood-based SE for MRL
compute_mrl_with_se <- function(fit) {
  time <- fit$time
  surv_probs <- fit$surv
  strata <- rep(names(fit$strata), fit$strata)
  
  df <- data.frame(time = time, surv = surv_probs, strata = strata) %>%
    group_by(strata) %>%
    arrange(time) %>%
    mutate(
      delta_t = lead(time, default = max(time)) - time,
      area = surv * delta_t,
      mrl = rev(cumsum(rev(area))),
      var = c(rev(cumsum(rev((delta_t^2) * (surv * (1 - surv) / sum(surv)))))),  # crude delta approx
      se = sqrt(var),
      lower = mrl - 1.96 * se,
      upper = mrl + 1.96 * se
    )
  return(df %>% select(time, mrl, se, lower, upper, strata))
}

mrl_df <- compute_mrl_with_se(fit)

# Split groups
mrl_ii <- filter(mrl_df, strata == "pathologic_stage=Stage II")
mrl_iii <- filter(mrl_df, strata == "pathologic_stage=Stage III")

# Shared time grid
max_common_time <- min(max(mrl_ii$time), max(mrl_iii$time))
shared_time <- sort(unique(c(mrl_ii$time, mrl_iii$time)))
shared_time <- shared_time[shared_time <= max_common_time]

# Step functions for each group
step_mrl_ii <- stepfun(mrl_ii$time, c(mrl_ii$mrl, tail(mrl_ii$mrl, 1)))
step_se_ii  <- stepfun(mrl_ii$time, c(mrl_ii$se, tail(mrl_ii$se, 1)))

step_mrl_iii <- stepfun(mrl_iii$time, c(mrl_iii$mrl, tail(mrl_iii$mrl, 1)))
step_se_iii  <- stepfun(mrl_iii$time, c(mrl_iii$se, tail(mrl_iii$se, 1)))

# Create aligned data
aligned_df <- data.frame(
  time = shared_time,
  MRL_II = step_mrl_ii(shared_time),
  SE_II = step_se_ii(shared_time),
  MRL_III = step_mrl_iii(shared_time),
  SE_III = step_se_iii(shared_time)
)

# Compute difference and CI
aligned_df <- aligned_df %>%
  mutate(
    MRL_Diff = MRL_II - MRL_III,
    SE_Diff = sqrt(SE_II^2 + SE_III^2),
    Lower_CI = MRL_Diff - 1.96 * SE_Diff,
    Upper_CI = MRL_Diff + 1.96 * SE_Diff
  )

# Plot with 95% CI ribbon
ggplot(aligned_df, aes(x = time, y = MRL_Diff)) +
  geom_step(color = "steelblue", size = 1.2) +
  geom_ribbon(aes(ymin = Lower_CI, ymax = Upper_CI), fill = "steelblue", alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.8) +
  labs(
    title = "Mean Residual Life Difference with 95% CI: Group A vs B ",
    x = "Time in Years",
    y = "MRL Difference Group A-B"
  ) +
  theme_minimal()

```

